A Method for Regularisation of Evolutionary Polynomial Regressions

### 2013-12-17 tests_Housing.Rmd
added tuned algs

>>>>

########################################################
# Tuning SVM & RPART

rsme <- function(v1, v2=c(0)) {
  sqrt(mean((v1-v2)^2))
}

library(e1071)

head(my.data,8)

n.runs <- 25
train.p.size <- 0.7 # percentage of training set

svm.error <- rep(NA,n.runs)
rpart.error <- rep(NA,n.runs)

for (i in 1:n.runs) {
  inTrain   <- sample(1:nrow(my.data), train.p.size * nrow(my.data))
  train.set <- my.data[inTrain,]
  test.set  <- my.data[-inTrain,]

  svm.tune <- tune.svm(y ~ ., data=train.set)
  svm.pred  <- predict(svm.tune$best.model, test.set[,-ncol(test.set)]) 
  svm.error[i] <- rsme(svm.pred,test.set[,ncol(test.set)])

  rpart.tune <- tune.rpart(y ~ ., data=train.set)
  rpart.pred  <- predict(rpart.tune$best.model, test.set[,-ncol(test.set)]) 
  rpart.error[i] <- rsme(rpart.pred,test.set[,ncol(test.set)])

  cat(i)
}

summary(svm.error)
summary(rpart.error)

### 2013-12-17 tests_Kinematics.Rmd
added tuned algs

>>>>

########################################################
# Tuning SVM & RPART

rsme <- function(v1, v2=c(0)) {
  sqrt(mean((v1-v2)^2))
}

library(e1071)

head(my.data,8)

n.runs <- 25
train.p.size <- 0.7 # percentage of training set

svm.error <- rep(NA,n.runs)
rpart.error <- rep(NA,n.runs)

for (i in 1:n.runs) {
  inTrain   <- sample(1:nrow(my.data), train.p.size * nrow(my.data))
  train.set <- my.data[inTrain,]
  test.set  <- my.data[-inTrain,]

  svm.tune <- tune.svm(y ~ ., data=train.set)
  svm.pred  <- predict(svm.tune$best.model, test.set[,-ncol(test.set)]) 
  svm.error[i] <- rsme(svm.pred,test.set[,ncol(test.set)])

  rpart.tune <- tune.rpart(y ~ ., data=train.set)
  rpart.pred  <- predict(rpart.tune$best.model, test.set[,-ncol(test.set)]) 
  rpart.error[i] <- rsme(rpart.pred,test.set[,ncol(test.set)])
}

summary(svm.error)
summary(rpart.error)

### 2013-12-17 texts/submission_02-asc/paper_spfga-review.tex
added tuned algs

%
%	ACRONYMS
%
\usepackage{acronym}
%
\acrodef{GA}{Genetic algorithms}
\acrodef{rmse}[error]{root-mean-square error}
\acrodef{GAPoly}[\textsc{GAPoly}]{Genetic Algorithms for Polynomials}
\acrodef{GAPolyreg}[\textsc{GAPoly.reg}]{Genetic Algorithms for Polynomials with Regularization}
\acrodef{EPR}{Evolutionary Polynomial Regression}
\acrodef{EPRR}[EPR$^2$]{Evolutionary Polynomial Regression with Regularisation}
%

@@

\address[ue]{Dept. Informática, Universidade de Évora, Rua Romão Ramalho 58, 7000-671 Évora}
\address[fcul]{Dept. Informática, Faculdade de Ciências da Universidade de Lisboa, Campo Grande 1749-016 Lisboa}
\address[labmag]{Laboratory of Agent Modelling (LabMAg)}
%
%\date{[{\sc draft} \today]}
%
\begin{abstract}
While many applications require models that have no acceptable linear approximation the simpler nonlinear models are defined by polynomials. The use of genetic algorithms to find polynomial models from data is known as \ac{EPR}.
%
This paper introduces \ac{EPRR}, an algorithm that extends the \ac{EPR} method and describes a set of experiences on common datasets that compare both flavours of \ac{EPR} and other methods including Linear Regression, Regression Trees and Support Vector Regression.

The empiric conclusion of those experiments is that \ac{EPRR} achieves better fitting than the other non-ensemble methods and shorter computation time than plain \ac{EPR}.
\end{abstract}
%
\begin{keyword}
evolutionary polynomial regression \sep regularisation \sep feature extraction \sep dimensionality reduction 
\end{keyword}
\end{frontmatter}
%
%	Line numbers for revision
%
%
%\linenumbers
%
\section{Introduction}

With notable exceptions (\emph{e.g.} neural networks) machine learning regression techniques produce linear models. The linearity assumption has many advantages including reduced computational complexity and strong the\-o\-re\-ti\-cal framework. However nonlinearity is unavoidable in many application scenarios, specially those with phase transitions or feedback loops, so common in engineering, ecology, cybernetics and other areas. The kernel trick in \acp{SVM} (\cite{scholkopf1997kernel, liang2012eigen, Bao:2013aa}) alleviates this problem by allowing special non-linear transformations of the feature-space. The condition such transformations must meet is known as the \emph{kernel trick}, $k\at{x,x'} = \left< \varphi\at{x}, \varphi\at{x'} \right>$, where $\varphi$ is the feature-space transformation and $\left<\cdot,\cdot\right>$ denotes inner product. The ``trick'' consists on computing the kernel $k\at{x,x'}$ while avoiding the computation of the inner product and the transformations $\varphi\at{x}, \varphi\at{x'}$. A special case of polynomial transformation, the \emph{polynomial kernel}, $k\at{x,x'} = \left<x,x'\right>^d$ is commonly used in regression and classification tasks with \acp{SVM}. However general polynomial transformations do not verify the kernel trick.

@@

This paper describes an extension of the general \ac{EPR} method to find a \emph{regularized} polynomial regression of a given dataset. The optimal regression results from a cost function that accounts for both the \ac{rmse} and a regularization factor to avoid over-fitting. Another important change is in the coding mechanism which corresponds to two new features: (a) inclusion of the linear terms in the polynomial expression, and (b) the activation/deactivation of monoids during evolution to allow the increase/decrease of polynomial expressions.

@@

The next section describes the details of our method and is followed by a presentation of some performance results. The last section draws some conclusions and points future research tasks.

@@

The second problem is treated in the \ac{GA} setting: Let $\mathcal{D}$ be a dataset as above and $Q$ a set of polynomials. For each polynomial $p\in Q$ let $\mathcal{P}_p$ be the set of monomials in $p$ (without the coefficients) and compute the fitness
%
\begin{eqnarray*}
\phi_p &=& \min_\Theta J_{\textrm{fit}}\left(\Theta;\mathcal{P}_p,\mathcal{D}\right)
\end{eqnarray*}
%
by solving the first problem. With a fitness of every instance, a \ac{GA} will apply genetic operators (usually mutation and crossover) to evolve the population $Q$ until a reasonable approximation of a local minimum is found. 
%
\begin{algorithm}[tb]
\begin{algorithmic}
\Function{GAPoly}{$D,pop_0,\epsilon$}
	\State $pop \gets pop_0;\: err \gets 1.0+\epsilon$
	\While{$err > \epsilon \land \neg\Call{limit}{iterations}$}
		\State $pop \gets \Call{IterateGA}{pop}$
		\State $pop \gets \Call{Sort}{pop, key = J}$\Comment{Sort population by regression error}
		\State $err \gets J\left(\Call{First}{pop}\right)$
	\EndWhile
	\State\Return $\Call{First}{pop}$
\EndFunction
\end{algorithmic}
\caption{\ac{GAPoly} uses linear regression to find monomial coefficients that minimize the \ac{rmse} over a dataset and \acp{GA} to explore the space of polynomials. The role of the linear regression step is to produce a fitting error that sorts the population. At exit the \ac{rmse} of the fittest instance is bounded by $\epsilon$ or the maximum number of allowed iterations.}\label{alg:gapoly}
\end{algorithm} 
%
Notice that the properties of \acp{GA} and linear regression entail that
the composition of \acp{GA} with linear regression, as defined in Algorithm \ref{alg:gapoly}, converges to a polynomial that is a local minimum of the fitness function, encapsulated in the \ac{rmse} function $J_{\textrm{fit}}$.
%

@@

Let's see an example: consider polynomial $x_1^3 x_3 + x_3^7 + x_1 x_2$ with $k = 4, m = 3$ and $d=3$. One possible encoding would be:
\begin{center}
\texttt{110 - 011,000,001 ; 000,000,111 ; 001,001,000 ; 110,010,101}
\end{center}

(for reading purposes the semicolons separate monomials, the commas separate variables)

The first three bits inform that the second and third monomials are active while the fourth is not (as said, the first monomial is always active). This last monomial does not enter neither in the polynomial regression nor in the \ac{rmse} (fitness) evaluation. However, it acts as a kind of junk DNA, becoming active when, in a future mutation or crossover, the third bit of the entire sequence flips from~0 to~1.

@@

In the experiments that led to this article, our tests concluded that the inclusion of the linear terms achieved a significant reduction of the error rates. So, considering the previous example, the final polynomial would be $x_1^3 x_3 + x_3^7 + x_1 x_2 + x_1 + x_2 + x_3$.

@@

This \ac{rmse} function tends to prefer more complex polynomials, namely in the number of monomials which provides the regression algorithm for more fitting possibilities. One way to balance this is to provide a regularization term into the \ac{rmse} function. Our proposal is to include a multiplicative factor proportional to the number of monomials. Thus, the error from equation \ref{eq:rmse} defines
\begin{eqnarray}
J_{\textrm{reg}}\left(\Theta;\mathcal{P},\mathcal{D}\right) &=& \lambda^{k} J_{\textrm{fit}}\left(\Theta;\mathcal{P},\mathcal{D}\right)\label{eq:rmse-reg}
\end{eqnarray}
%
where $k$ is the number of monomials in the polynomial. When $\lambda > 1$ polynomials with more monomials are penalized.

Somewhat unexpectedly after some experiences it was found that lower values for $\lambda$ provide better, even if marginal, results. Figure~\ref{Abalone_dataset_lambdas} shows regularization results for the Abalone and Auto MPG datasets with ten runs for each $\lambda$. The following section  includes information about these datasets.

%
\begin{figure*}[tb]\begin{center}
\begin{tabular}{cc}
\includegraphics[width=0.49\textwidth]{figure_1a.pdf}
%
% RIGHT PANE
&
%
%
\includegraphics[width=0.49\textwidth]{figure_1b.pdf}
\end{tabular}

@@

The typical inflection point lies around $\lambda = 0.8$. The dataset results for applying the proposed regression method use the regularization parameter with this value.

@@

The results were found using \texttt{R} programming language~\citep{R}\footnote{The datasets and the \texttt{R} code used to produce the results and plots in this paper are available online at \url{https://github.com/jpneto/GenAlgPoly}.}. To compare this paper's proposed algorithm we applied the exact same train and test samples using several well-known learning algorithms for regression, namely: the classic \ac{GAPoly}, Linear Regression, Support Vector Machines~\citep{Meyer12}, Regression Trees~\citep{Therneau13} and Conditional Inference Trees~\citep{Hothorn06, Strobl07, Strobl08}.

@@

@@

In the next datasets, the population of the genetic algorithm had size~$n=250$ with 100~iterations for each run.

@@

The \ac{GA} quickly proceeds in the first~50 to~100 generations to reasonable error rates. Then, it proceeds slower achieving best solutions with marginal error reduction. Since the entire learning process takes some time, in the current R implementation, placing a limit between 50 to 100 generations already achieves good results, relative to higher iteration values. Figure~\ref{Abalone_fitnessProgress} shows a typical error evolution for the dataset Abalone given two different values for $\lambda$.

@@

Of all the non-ensemble regression methods considered, the proposed method is the one that shows, with one exception, the lowest error. Only Random Forest outperforms \ac{GAPoly} systematically (it also outperforms all the other regression algorithms). The single non-ensemble exception --- besides the artificial dataset that uses a straightforward polynomial relation --- is the Auto MPG dataset where \ac{GAPoly} has comparable results. This is evidence that applying standard genetic operators for polynomial model searching is a viable tool for regression purposes.

For complexity considerations \ac{GAPoly} demands some processing time. On a current quad-core computer, processing the Kinematics dataset (with 8k~observations) takes approximately 5~minutes. The processing time can probably be speeded by one to two orders in magnitude if the process is implemented in a low level programming language like \texttt{C++}. However, speed optimization was not the focus of this article.

A cross-validation procedure can be implemented to refine the appropriate parameter values to achieve better errors. Namely, the regularization parameter, $\lambda$, can be tested with several values, instead of being fixed at $0.8$. Other parameters like mutation chance or the amount of elitism could also be tested. However, these type of tests need a low-level, fast implementation of \ac{GAPoly}.

>>>>

%
%	ACRONYMS
%
\usepackage{acronym}
%
\acrodef{GA}{Genetic algorithms}
\acrodef{rmse}[error]{root-mean-square error}
%\acrodef{GAPoly}[\textsc{GAPoly}]{Genetic Algorithms for Polynomials}
%\acrodef{GAPolyreg}[\textsc{GAPoly.reg}]{Genetic Algorithms for Polynomials with Regularization}

\acrodef{GAPoly}[EPR]{Evolutionary Polynomial Regression}
\acrodef{GAPolyreg}[EPR$^2$]{Evolutionary Polynomial Regression with Regularisation}

\acrodef{EPR}{Evolutionary Polynomial Regression}
\acrodef{EPRR}[EPR$^2$]{Evolutionary Polynomial Regression with Regularisation}

\acrodef{SVM}[SVM]{Support Vector Machines}
%

@@

\address[ue]{Dept. Informática, Universidade de Évora, Rua Romão Ramalho 58, 7000-671 Évora}
\address[fcul]{Dept. Informática, Faculdade de Ciências da Universidade de Lisboa, Campo Grande 1749-016 Lisboa}
\address[labmag]{Laboratory of Agent Modelling (LabMAg)}
%
%\date{[{\sc draft} \today]}
%
\begin{abstract}
While many applications require models that have no acceptable linear approximation, the simpler nonlinear models are defined by polynomials. The use of genetic algorithms to find polynomial models from data is known as \ac{EPR}.
%
This paper introduces \ac{EPRR}, an algorithm that extends the \ac{EPR} method and describes a set of experiences on common datasets that compare both flavours of \ac{EPR} and other methods including Linear Regression, Regression Trees and Support Vector Regression.

The empiric conclusion of those experiments is that \ac{EPRR} is able to achieve better fitting than other non-ensemble methods and it has shorter computation time than plain \ac{EPR}.
\end{abstract}
%
\begin{keyword}
evolutionary polynomial regression \sep regularisation \sep feature extraction \sep dimensionality reduction 
\end{keyword}
\end{frontmatter}
%
%	Line numbers for revision
%
%
%\linenumbers
%
\section{Introduction}

With notable exceptions (\emph{e.g.} neural networks) machine learning regression techniques produce linear models. The linearity assumption has many advantages including reduced computational complexity and strong the\-o\-re\-ti\-cal framework. However nonlinearity is unavoidable in many application scenarios, specially those with phase transitions or feedback loops, so common in engineering, ecology, cybernetics and other areas. The kernel trick in \ac{SVM} (\cite{scholkopf1997kernel, liang2012eigen, Bao:2013aa}) alleviates this problem by allowing special non-linear transformations of the feature-space. The condition such transformations must meet is known as the \emph{kernel trick}, $k\at{x,x'} = \left< \varphi\at{x}, \varphi\at{x'} \right>$, where $\varphi$ is the feature-space transformation and $\left<\cdot,\cdot\right>$ denotes inner product. The ``trick'' consists on computing the kernel $k\at{x,x'}$ while avoiding the computation of the inner product and the transformations $\varphi\at{x}, \varphi\at{x'}$. A special case of polynomial transformation, the \emph{polynomial kernel}, $k\at{x,x'} = \left<x,x'\right>^d$ is commonly used in regression and classification tasks with \acp{SVM}. However general polynomial transformations do not verify the kernel trick.

@@

This paper describes an extension of the general \ac{EPR} method to find a \emph{regularized} polynomial regression of a given dataset. The optimal regression results from a cost function that accounts for both the \ac{rmse} and a regularization factor to avoid over-fitting. Another important change is in the coding mechanism which introduces two new features: (a) inclusion of the linear terms in the polynomial expression, and (b) the activation/deactivation of monoids during evolution to allow the increase/decrease of polynomial expressions.

@@

The next section describes the method's details and is followed by a presentation of some performance results. The last section draws some conclusions and points future research tasks.

@@

The second problem is treated in the \ac{GA} setting: Let $\mathcal{D}$ be a dataset as above and $Q$ a set of polynomials. For each polynomial $p\in Q$ let $\mathcal{P}_p$ be the set of monomials in $p$ (without the coefficients) and compute the fitness
%
\begin{eqnarray*}
\phi_p &=& \min_\Theta J_{\textrm{fit}}\left(\Theta;\mathcal{P}_p,\mathcal{D}\right)
\end{eqnarray*}
%
by solving the first problem. With a fitness of every instance, a \ac{GA} will apply genetic operators (usually mutation and crossover) to evolve the population $Q$ until a reasonable approximation of a local minimum is found. 
%
\begin{algorithm}[tb]
\begin{algorithmic}
\Function{EPR}{$D,pop_0,\epsilon$}
	\State $pop \gets pop_0;\: err \gets 1.0+\epsilon$
	\While{$err > \epsilon \land \neg\Call{limit}{iterations}$}
		\State $pop \gets \Call{IterateGA}{pop}$
		\State $pop \gets \Call{Sort}{pop, key = J}$\Comment{Sort population by regression error}
		\State $err \gets J\left(\Call{First}{pop}\right)$
	\EndWhile
	\State\Return $\Call{First}{pop}$
\EndFunction
\end{algorithmic}
\caption{\ac{GAPoly} uses linear regression to find monomial coefficients that minimize the \ac{rmse} over a dataset and \acp{GA} to explore the space of polynomials. The role of the linear regression step is to produce a fitting error that sorts the population. At exit the \ac{rmse} of the fittest instance is bounded by $\epsilon$ or the maximum number of allowed iterations.}\label{alg:gapoly}
\end{algorithm} 
%
Notice that the properties of \acp{GA} and linear regression entail that
the composition of \acp{GA} with linear regression, as defined in Algorithm \ref{alg:gapoly}, converges to a polynomial that is a local minimum of the fitness function, encapsulated in the \ac{rmse} function $J_{\textrm{fit}}$.
%

@@

A short example: consider polynomial $x_1^3 x_3 + x_3^7 + x_1 x_2$ with $k = 4, m = 3$ and $d=3$. A possible encoding would be (for reading purposes the semicolons separate monomials, the commas separate variables):
\begin{center}
\texttt{110 - 011,000,001 ; 000,000,111 ; 001,001,000 ; 110,010,101}
\end{center}

The first three bits inform that the second and third monomials are active while the fourth is not (the first monomial is always active). This fourth monomial does not enter neither in the polynomial regression nor in the \ac{rmse} (fitness) evaluation. However, it acts as a kind of junk DNA, becoming active when, in a future mutation or crossover, the third bit of the entire sequence flips from~0 to~1.

@@

In the experiments that led to this article, the tests showed that the inclusion of the linear terms achieved a significant reduction of error rates. So, considering the previous example, the final polynomial would be $x_1^3 x_3 + x_3^7 + x_1 x_2 + x_1 + x_2 + x_3$. The next section discusses this feature.

@@

This \ac{rmse} function tends to prefer more complex polynomials, namely in the number of monomials which provides the regression algorithm for more fitting possibilities. One way to balance this is to provide a regularization term into the \ac{rmse} function. The article's proposal is to include a multiplicative factor proportional to the number of monomials. Thus, the error from equation \ref{eq:rmse} defines
\begin{eqnarray}
J_{\textrm{reg}}\left(\Theta;\mathcal{P},\mathcal{D}\right) &=& \lambda^{k} J_{\textrm{fit}}\left(\Theta;\mathcal{P},\mathcal{D}\right)\label{eq:rmse-reg}
\end{eqnarray}
%
where $k$ is the number of monomials in the polynomial. When $\lambda > 1$ polynomials with more monomials are penalized.

\begin{figure*}[tb]\begin{center}
\begin{tabular}{cc}
\includegraphics[width=0.49\textwidth]{figure_1a.pdf}
%
% RIGHT PANE
&
%
%
\includegraphics[width=0.49\textwidth]{figure_1b.pdf}
\end{tabular}

@@

Figure~\ref{Abalone_dataset_lambdas} shows regularization results for the Abalone and Auto MPG datasets with ten runs for each $\lambda$ (the following section  includes information about these datasets). The typical inflection point lies around $\lambda = 0.8$. This somewhat paradoxical result -- since regularization tries to minimize the model's complexity in order to prevent overfitting -- is somewhat justified by the inclusion of the linear terms mentioned in the previous section. The linear terms `pushes' the model to linearity while a regularization term $\lambda < 1$ tries to add non-linearity terms. Within this tension, the algorithm is able to reduce the overall error, compared to the typical implementation \ac{EPR} found in the literature.

@@

The results were found using \texttt{R} programming language~\citep{R}\footnote{The datasets and the \texttt{R} code used to produce the results and plots in this paper are available online at \url{https://github.com/jpneto/GenAlgPoly}.}. To compare this paper's proposed algorithm we applied the exact same train and test samples using several well-known learning algorithms for regression, namely: the classic \ac{GAPoly}, Linear Regression, Support Vector Machines~\citep{Meyer12}, Regression Trees~\citep{Therneau13} and Conditional Inference Trees~\citep{Hothorn06, Strobl07, Strobl08}. To achieve better error results, the \ac{SVM} and Regression Tree algorithms had their parameters tuned.

@@

In the next datasets, the population of the genetic algorithm had size~$n=300$ with 100~iterations for each run. \ac{GAPolyreg} was run with regularization parameter $lambda = 0.8$.

@@

@@

The \ac{GA} quickly proceeds in the first~50 to~100 generations to reasonable error rates. Then, it proceeds slower achieving best solutions with marginal error reduction. Since the entire learning process takes some time, in the current R implementation, placing a limit between 50 to 100 generations already achieves good results, relative to higher iteration values. Figure~\ref{Abalone_fitnessProgress} shows a typical error evolution for the dataset Abalone given a regularized version ($\lambda = 0.975$) and a non-regularized ($\lambda = 1.0$).

@@

% Of all the regression methods considered, the proposed method is the one that shows, with one exception, the lowest error. Only Random Forest outperforms \ac{GAPoly} systematically (it also outperforms all the other regression algorithms). The single non-ensemble exception --- besides the artificial dataset that uses a straightforward polynomial relation --- is the Auto MPG dataset where \ac{GAPoly} has comparable results. This is evidence that applying standard genetic operators for polynomial model searching is a viable tool for regression purposes.

Of the regression methods considered, \ac{SVM} achieved the best results in 3 out of 4 datasets. However, \ac{SVM} was tuned for each particular dataset while \ac{GAPolyreg} was executed with its parameters preset. Even so, \ac{GAPolyreg} had the best error in the Abalone dataset, and achieved competitive results in the Housing and Auto MPG datasets.

Comparing \ac{GAPoly} and \ac{GAPolyreg} -- which is the main article's topic -- the regularized version achieved much better results at Abalone and especially Kinematics. The Housing improved errors when compared with \ac{GAPoly} in a difference in means, resulted in a 95\% HDI (Highest Density Interval) equal to [0.001, 0.119] which, while borderline, achieves statistical significance. Only in the Auto MOG dataset \ac{GAPoly} achieved better results, even if not that different from \ac{GAPolyreg}.

For complexity considerations \ac{GAPoly} and \ac{GAPolyreg} demands some processing time. On a quad-core computer, processing the Kinematics dataset (with 8k~observations) takes approximately 5~minutes. The processing time can probably be speeded between one to two orders in magnitude if the process is implemented in a low level programming language like \texttt{C++}. However, speed optimization was not the focus of this article.

A cross-validation procedure can be implemented to refine the appropriate parameter values to achieve better errors. Namely, the regularization parameter, $\lambda$, can be tested with several values, instead of being fixed at $0.8$. Other parameters like mutation chance or the amount of elitism could also be tested. However, these type of tests need a low-level, fast implementation of \ac{GAPoly} and were not considered.
